A Typed Programmatic Interface to Contracts on the Blockchain

### 2021-02-17 papers/execution-model/model.tex
explain typing

\begin{figure}[tp]
\begin{align*}
  \CONSTANT & ::= \INT \mid \FIX
              \mid \OPH \mid \PUH \mid\PUK \mid \CODE \mid
              \NTEZ \mid \SUNIT \mid \FALSE \mid \TRUE \\
  \STATUS & ::= \STATUSPENDING \mid \STATUSINCLUDING (\INT) \mid
            \STATUSTIMEOUT \\
  \ERROR &::= \ERRPRG \mid \ERRBAL \mid \ERRCOUNT \mid \ERRFEE \mid
           \ERRPUK \mid \ERRPUH \mid \ERRARG \mid \ERRINIT\\
	\EXPR & :: =  \CONSTANT \mid \STATUS \mid \ERROR \mid \VARIABLE \mid \lambda \VARIABLE. \EXPR\ 
	\mid \EXPR\EXPR  %\mid \FST\ \EXPR \mid \SND\ \EXPR 	%\mid \NI \EI
  \mid \EXPR\ \PLUS\ \EXPR\ \mid
  % \EXPR\ \MINUS\ \EXPR\ \mid
  \EXPR\ \EQUAL\ \EXPR\
  % \mid  \EXPR\ \LESS\ \EXPR
  \mid \EXPR\ \AND\ \EXPR\ \mid \EXPR\ \OR\ \EXPR\ \mid \NOT\ \EXPR
  \\& \mid (\EXPR,\EXPR)\mid \NIL\ \mid \CONS\ \EXPR\ \EXPR \mid \LEFT\ \EXPR \mid\RIGHT\
  \EXPR \mid \SOME\ \EXPR \mid \NONE
  \mid \MATCH\ \EXPR\ \WITH\ \PATTERN\to\EXPR \dots 
  \\& \mid \RAISE\ \EXPR \mid \TRY\ \EXPR\ \EXCEPT\ \EXPR
  \\& \mid \TRANSFER[\EXPR]\EXPR\EXPR\EXPR\EXPR
  \\& \mid \ORIGINATE\EXPR\EXPR\EXPR\EXPR\EXPR
  \\&\mid \QOP\  \EXPR \mid
  \CONTRACT\ \TYPE\ \EXPR \mid \TOSTRING\
  \EXPR \mid \FROMSTRING\ \TYPE\ \EXPR
  \\
  \PATTERN &::= \VARIABLE \mid (\PATTERN, \PATTERN) \mid \NIL \mid \CONS\ \PATTERN\ \PATTERN \mid \LEFT\ \PATTERN \mid\RIGHT\
             \PATTERN \mid \SOME\ \PATTERN \mid \NONE \\
  &\mid \FALSE \mid \TRUE \mid \STATUS \mid \ERROR\\
  \QOP & ::= balance \mid status \mid storage \mid code \mid counter
         \mid type
  \\[2ex]
  \TYPE, \TYPEU & ::=
                  \TADDR \mid \TOPH \mid
                  \TCONTRACT\ \TYPE \mid
                  % \TPUH\ \TYPE \mid
                  \TPUK \mid
                  \TCODE\ \TYPE\ \TYPEU \mid \TTEZ \mid
          \TSTATUS \mid \TEXCEPTION \\
  & \mid \TINT \mid \TUNIT \mid \TBOOL \mid \TSTRING \mid \TYPE\to\TYPEU \mid \TPAIR\ \TYPE\ \TYPEU \mid \TLIST\ \TYPE
    \mid \TSUM\ \TYPE\ \TYPEU \mid \TOPTION\ \TYPE 
\end{align*}
  \caption{Syntax of expressions, \EXPR, and types, \TYPE}
  \label{fig:syntax-expressions}
\end{figure}
All queries and operations are started by closed expressions that run on the
local node. Figure~\ref{fig:syntax-expressions} defines the syntax of
expressions, which describe the usual constructs of call-by-value PCF
with sum, product, list, and option types and exceptions. Moreover, there are
primitive types and constants \CONSTANT\ to support blockchain
operations, as well as several exceptional values collected in \ERROR.
Pattern matching is the only means to decompose values, cf.\ \PATTERN.
There is syntax to initiate transfers and to originate contracts as
well as for the queries. Finally, the three constructs 
\CONTRACT\ \TYPE\ \EXPR, \TOSTRING\  \EXPR, and \FROMSTRING\ \TYPE\
\EXPR\ are casts between untyped and typed representations of
contracts and their parameters.

@@

\begin{figure}[tp]
  \begin{align*}
    \OPH &=  \TRANSFER[\STRING]\NTEZ\PUK\PUH\MTEZ \\
    \\
         & \MATCH\ {status\ \OPH}\ \WITH \\
         & \mid \STATUSPENDING \to \textrm{not yet included, continue waiting} \\
         & \mid \STATUSINCLUDING (n) \to \textrm{included with $n$
           blocks confirmation}\\
         & \mid \STATUSTIMEOUT \to \textrm{lost, try again}
  \end{align*}
  \caption{Pseudocode example}
  \label{fig:pseudocode-example}
\end{figure}
Figure~\ref{fig:pseudocode-example} shows an expression that invokes a
contract at \PUH\ with argument \STRING.

Subtyping
\begin{mathpar}
  \inferrule{}{\TPUH \SubType \TADDR}

@@

  \inferrule{\TYPEU' \SubType \TYPE' \\ \TYPE \SubType \TYPEU}{
    \TYPE' \to \TYPE \SubType \TYPEU' \to \TYPEU}
\end{mathpar}

Figure~\ref{fig:typing-expressions} presents selected typing rules for
expressions. We rely on an external typing judgment
$\JTypeCode\CODE\TYPE$ for the contract language, which we leave
unspecified, and $\JTypeValue\STRING\TYPE$ for serialized values as
stored on the blockchain. The latter judgment states $\STRING$ is
a string parseable as a value of type $\TYPE$.
\begin{figure}[tp]
  \begin{mathpar}
    \inferrule{}{\JTypeExpr\TEnv\INT\TINT}

>>>>

\begin{figure}[tp]
\begin{align*}
  \CONSTANT & ::= \INT \mid \FIX
              \mid \OPH \mid \PUH \mid\PUK \mid \CODE \mid
              \NTEZ \mid \SUNIT \mid \FALSE \mid \TRUE \\
  \STATUS & ::= \STATUSPENDING \mid \STATUSINCLUDING (\INT) \mid
            \STATUSTIMEOUT \\
  \ERROR &::= \ERRPRG \mid \ERRBAL \mid \ERRCOUNT \mid \ERRFEE \mid
           \ERRPUK \mid \ERRPUH \mid \ERRARG \mid \ERRINIT\\
	\EXPR & :: =  \CONSTANT \mid \STATUS \mid \ERROR \mid \VARIABLE \mid \lambda \VARIABLE. \EXPR\ 
	\mid \EXPR\EXPR  %\mid \FST\ \EXPR \mid \SND\ \EXPR 	%\mid \NI \EI
  \mid \EXPR\ \PLUS\ \EXPR\ \mid
  % \EXPR\ \MINUS\ \EXPR\ \mid
  \EXPR\ \EQUAL\ \EXPR\
  % \mid  \EXPR\ \LESS\ \EXPR
  \mid \EXPR\ \AND\ \EXPR\ \mid \EXPR\ \OR\ \EXPR\ \mid \NOT\ \EXPR
  \\& \mid (\EXPR,\EXPR)\mid \NIL\ \mid \CONS\ \EXPR\ \EXPR \mid \LEFT\ \EXPR \mid\RIGHT\
  \EXPR \mid \SOME\ \EXPR \mid \NONE
  \mid \MATCH\ \EXPR\ \WITH\ \PATTERN\to\EXPR \dots 
  \\& \mid \RAISE\ \EXPR \mid \TRY\ \EXPR\ \EXCEPT\ \EXPR
  \\& \mid \TRANSFER[\EXPR]\EXPR\EXPR\EXPR\EXPR
  \\& \mid \ORIGINATE\EXPR\EXPR\EXPR\EXPR\EXPR
  \\&\mid \QOP\  \EXPR \mid
  \CONTRACT\ \TYPE\ \EXPR \mid \TOSTRING\
  \EXPR \mid \FROMSTRING\ \TYPE\ \EXPR
  \\
  \PATTERN &::= \VARIABLE \mid (\PATTERN, \PATTERN) \mid \NIL \mid \CONS\ \PATTERN\ \PATTERN \mid \LEFT\ \PATTERN \mid\RIGHT\
             \PATTERN \mid \SOME\ \PATTERN \mid \NONE \\
  &\mid \FALSE \mid \TRUE \mid \STATUS \mid \ERROR\\
  \QOP & ::= balance \mid status \mid storage \mid code \mid counter
         \mid type
  \\[2ex]
  \TYPE, \TYPEU & ::=
                  \TPUH \mid
                  \TPUK \mid
                  \TADDR \mid 
                  \TCONTRACT\ \TYPE \mid
                  \TCODE\ \TYPE\ \TYPEU \mid \TTEZ \mid
                  \TOPH \mid
                  \TSTATUS \mid \TEXCEPTION \\
  & \mid \TINT \mid \TUNIT \mid \TBOOL \mid \TSTRING \mid \TYPE\to\TYPEU \mid \TPAIR\ \TYPE\ \TYPEU \mid \TLIST\ \TYPE
    \mid \TSUM\ \TYPE\ \TYPEU \mid \TOPTION\ \TYPE 
\end{align*}
  \caption{Syntax of expressions, \EXPR, and types, \TYPE}
  \label{fig:syntax-expressions}
\end{figure}
All queries and operations are started by closed expressions that run on the
local node. Figure~\ref{fig:syntax-expressions} defines the syntax of
expressions, which describe the usual constructs of call-by-value PCF
with sum, product, list, and option types and exceptions. Moreover, there are
primitive types and constants \CONSTANT\ to support blockchain
operations, as well as several exceptional values collected in \ERROR.
Pattern matching is the only means to decompose values, cf.\ \PATTERN.
There is syntax to initiate transfers and to originate contracts as
well as for the queries. Finally, the three constructs 
\CONTRACT\ \TYPE\ \EXPR, \TOSTRING\  \EXPR, and \FROMSTRING\ \TYPE\
\EXPR\ are casts between untyped and typed representations of
contracts and their parameters.

Types (also in Figure~\ref{fig:syntax-expressions}) comprise a
collection of standard base types as well as 
functions, pairs, lists, sums, and option types. In addition, there
are domain specific types of public hashes $\TPUH$ and public keys $\TPUK$ subsumed by
a type of addresses $\TADDR$. $\TCONTRACT\ \TYPE$ is the type of
a contract with parameter type $\TYPE$. $\CODE\ \TYPE\ \TYPEU$
indicates a Michelson program with parameter type $\TYPE$ and storage
type $\TYPEU$. Tezos tokens have type $\TTEZ$. The type $\TOPH$ signifies operation hashes
returned by blockchain operations. Querying the status of an operation
returns a value of type $\TSTATUS$. Exceptions have type
$\TEXCEPTION$.

@@

\begin{figure}[tp]
  \begin{align*}
    \LET\ \OPH &=  \TRANSFER\NTEZ\PUK{\PUK'}\MTEZ \\
         & \MATCH\ \QRY\ \GETSTATUS\OPH\ \WITH \\
         & \mid \STATUSPENDING \to \textrm{not yet included, continue waiting} \\
         & \mid \STATUSINCLUDING (n) \to \textrm{included with $n$
           blocks confirmation}\\
         & \mid \STATUSTIMEOUT \to \textrm{lost operation, try again}
  \end{align*}
  \caption{Pseudocode example for a token transfer}
  \label{fig:pseudocode-example}
\end{figure}
Figure~\ref{fig:pseudocode-example} shows an expression that invokes a
contract at \PUH\ with argument \STRING.

Subtyping in Figure~\ref{fig:subtyping} etablishes that public hashes
and public keys can both stand for addresses. Moreover, a smart
contract with parameter type $\TYPE$ is represented by its public hash
at run time. Subtyping for function types has the usual co-/contravariant definition.
\begin{figure}[tp]
  \begin{mathpar}
    \inferrule{}{\TPUH \SubType \TADDR}

@@

    \inferrule{\TYPEU' \SubType \TYPE' \\ \TYPE \SubType \TYPEU}{
      \TYPE' \to \TYPE \SubType \TYPEU' \to \TYPEU}
  \end{mathpar}
  \caption{Subtyping}
  \label{fig:subtyping}
\end{figure}
Figure~\ref{fig:typing-expressions} presents selected typing rules for
expressions. We rely on an external typing judgment
$\JTypeCode\CODE\TYPE$ for the contract language, which we leave
unspecified, and $\JTypeValue\STRING\TYPE$ for serialized values as
stored on the blockchain. The latter judgment states $\STRING$ is
a string parseable as a value of type $\TYPE$.
\begin{figure}[tp]
  \begin{mathpar}
    \inferrule{}{\JTypeExpr\TEnv\INT\TINT}
